---
title: "HUDM - 5123 Final Project"
author: "Jon Zhang"
date: "2026-06-18"
output: html_document
---
###############################################################
# HUDM 5123 Final Project
#
# Replication of:
# Champion, Economides, & Chandler (2018)
#
# The efficacy of a brief app-based mindfulness intervention
# on psychosocial outcomes in healthy adults
#
# Author: Jonathan Zhang & Luyi Xu
# Date: 07/01/2026
###############################################################
# Load the Packages
```{r}
# Packages
library(httr)
library(readxl) # for reading .xlsx files
library(tidyverse) # using dylpr & recoding
library(janitor)
library(psych)
library(afex)
library(emmeans) # to compute estemated marginal means
library(effectsize) # to compute effect sizes
library(car)
library(rstatix)
```

# Uploading the Dataset
```{r}
# Define the RAW GitHub URL (As dataset is in .xlsx)
github_raw_url <- "https://github.com/ZeddyCraft/HUDM-5123-Final-Project/raw/refs/heads/main/Raw%20Data.xlsx"

# 3. Create a temporary local file path
temp_file <- tempfile(fileext = ".xlsx")

# 4. Download the file securely
GET(github_raw_url, write_disk(temp_file, overwrite = TRUE))

# 5. Read the Excel file into R
data <- read_excel(temp_file)

# View the first few rows of the data
head(data)
```

# Importing and Cleaning the Dataset
```{r}
# Clean variable names
data <- clean_names(data)

# Remove completely empty rows
data <- data %>%
  remove_empty(which = "rows")

# Create Participant ID
data <- data %>%
  mutate(id = 1:n())

# Move ID to the first column
data <- data %>%
  relocate(id)

# View the cleaned data
glimpse(data)
```
# Rename variables
```{r}
data <- data %>%
  rename(
    group = group_a_mindfulness_b_wait_list,
    gender = gender_1_male_2_female,
    
    swls_b = swls_baseline,
    swls_d10 = swls_day_10,
    swls_d30 = swls_day_30,
    
    pss_b = pss_baseline,
    pss_d10 = pss_day_10,
    pss_d30 = pss_day_30,
    
    rws_b = rws_baseline,
    rws_d10 = rws_day_10,
    rws_d30 = rws_day_30,

    sessions_10 = number_of_days_headspace_used_between_baseline_to_day_10,
    sessions_30 = number_of_days_headspace_used_between_day_10_and_day_30,

    enjoy_10 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_10_1_not_at_all_to_7_extremely,
    enjoy_30 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_30_1_not_at_all_to_7_extremely,

    difficulty_10 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_10_1_very_difficult_to_7_very_easy,
    difficulty_30 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_30_1_very_difficult_to_7_very_easy
  )

data <- data %>%
  mutate(
    group = factor(group,
                   levels = c("A", "B"),
                   labels = c("Mindfulness", "Wait-list")),

    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )
```

# Descriptive Statistics
```{r}
# Number of participants in each group
table(data$group)

# Percentage in each group
prop.table(table(data$group)) * 100

# Summary of age
describe(data$age)

# Summary by group
data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  )

# Gender distribution
table(data$group, data$gender)

# Gender percentages within each group
prop.table(table(data$group,
                 data$gender), margin = 1) * 100
```

```{r}
# Means and Standard Deviations
data %>%
  group_by(group) %>%
  summarise(
    # Satisfaction with Life
    SWLS_B_Mean = mean(swls_b, na.rm = TRUE),
    SWLS_B_SD   = sd(swls_b, na.rm = TRUE),

    SWLS_D10_Mean = mean(swls_d10, na.rm = TRUE),
    SWLS_D10_SD   = sd(swls_d10, na.rm = TRUE),

    SWLS_D30_Mean = mean(swls_d30, na.rm = TRUE),
    SWLS_D30_SD   = sd(swls_d30, na.rm = TRUE),

    # Perceived Stress
    PSS_B_Mean = mean(pss_b, na.rm = TRUE),
    PSS_B_SD   = sd(pss_b, na.rm = TRUE),

    PSS_D10_Mean = mean(pss_d10, na.rm = TRUE),
    PSS_D10_SD   = sd(pss_d10, na.rm = TRUE),

    PSS_D30_Mean = mean(pss_d30, na.rm = TRUE),
    PSS_D30_SD   = sd(pss_d30, na.rm = TRUE),

    # Resilience
    RWS_B_Mean = mean(rws_b, na.rm = TRUE),
    RWS_B_SD   = sd(rws_b, na.rm = TRUE),

    RWS_D10_Mean = mean(rws_d10, na.rm = TRUE),
    RWS_D10_SD   = sd(rws_d10, na.rm = TRUE),

    RWS_D30_Mean = mean(rws_d30, na.rm = TRUE),
    RWS_D30_SD   = sd(rws_d30, na.rm = TRUE)
  )
```

# Reshaping the Data (wide to Long)
```{r}
# Satisfaction With Life Scale
data_long <- data %>%
  select(id, group, swls_b, swls_d10, swls_d30) %>%
  pivot_longer(cols = starts_with("swls"), names_to = "time", values_to = "swls") %>%
  mutate(time = case_when(
         time == "swls_b"   ~ "Baseline",
         time == "swls_d10" ~ "Day 10",
         time == "swls_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Perceived Stress Scale (PSS)
pss_long <- data %>%
  select(id, group, pss_b, pss_d10, pss_d30) %>%
  pivot_longer(cols = starts_with("pss"), names_to = "time", values_to = "pss") %>%
  mutate(time = case_when(
         time == "pss_b" ~ "Baseline",
         time == "pss_d10" ~ "Day 10",
         time == "pss_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Resilience Scale (RWS)
rws_long <- data %>%
  select(id, group, rws_b, rws_d10, rws_d30) %>%
  pivot_longer(cols = starts_with("rws"), names_to = "time", values_to = "rws") %>%
  mutate(time = case_when(
         time == "rws_b" ~ "Baseline",
         time == "rws_d10" ~ "Day 10",
         time == "rws_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Check the dataset
head(rws_long)
glimpse(rws_long)
```

# Assumption Checking
```{r}
# Missing Data
## Missing values in each long dataset
colSums(is.na(data_long))
colSums(is.na(pss_long))
colSums(is.na(rws_long))

## Total missing observations
sum(is.na(data_long$swls))
sum(is.na(pss_long$pss))
sum(is.na(rws_long$rws))
```
```{r}
# Outliers (by Boxplots)
## SWLS
ggplot(data_long, aes(x = time, y = swls,fill = group)) +
  geom_boxplot() +
  labs(title = "SWLS by Time and Group", x = "Time", y = "SWLS Score") +
  theme_minimal()

## PSS
ggplot(pss_long, aes(x = time, y = pss, fill = group)) +
  geom_boxplot() +
  labs(title = "PSS by Time and Group", x = "Time", y = "PSS Score") +
  theme_minimal()

## RWS
ggplot(rws_long, aes(x = time, y = rws, fill = group)) +
  geom_boxplot() +
  labs(title = "RWS by Time and Group", x = "Time", y = "RWS Score") +
  theme_minimal()
```

```{r}
# Outliers (by Outlier Detection)
data_long %>%
  group_by(group, time) %>%
  identify_outliers(swls)

pss_long %>%
  group_by(group, time) %>%
  identify_outliers(pss)

rws_long %>%
  group_by(group, time) %>%
  identify_outliers(rws)
```

```{r}
# Normality
## SWLS
data_long %>%
  group_by(group, time) %>%
  shapiro_test(swls)

## PSS
pss_long %>%
  group_by(group, time) %>%
  shapiro_test(pss)

## RWS
rws_long %>%
  group_by(group, time) %>%
  shapiro_test(rws)
```

```{r}
## QQ Plots
## SWLS
ggplot(data_long, aes(sample = swls)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()

## PSS
ggplot(pss_long, aes(sample = pss)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()

## RWS
ggplot(rws_long, aes(sample = rws)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()
```

```{r}
# Homogeneity of Variance (Levene's Test) 
## SWLS
leveneTest(swls ~ group * time, data = data_long)

## PSS
leveneTest(pss ~ group * time,data = pss_long)

## RWS
leveneTest(rws ~ group * time, data = rws_long)
```
# Mixed Repeated-Measures ANOVA
```{r}
# Satisfaction With Life (SWLS)
swls_aov <- aov_ez(
  id = "id",
  dv = "swls",
  between = "group",
  within = "time",
  data = data_long,
  type = 3)

# Display ANOVA table
summary(swls_aov)

# Detailed ANOVA table
nice(swls_aov)
```

```{r}
# Perceived Stress (PSS)
pss_aov <- aov_ez(
  id = "id",
  dv = "pss",
  between = "group",
  within = "time",
  data = pss_long,
  type = 3)

# Display ANOVA table
summary(pss_aov)

# Detailed ANOVA table
nice(pss_aov)
```

```{r}
# Section 6.3: Mixed ANOVA - Resilience (RWS)
rws_aov <- aov_ez(
  id = "id",
  dv = "rws",
  between = "group",
  within = "time",
  data = rws_long,
  type = 3)

# Display ANOVA table
summary(rws_aov)

# Detailed ANOVA table
nice(rws_aov)
```

```{r}
# Effect Sizes
effectsize::eta_squared(swls_aov)
effectsize::eta_squared(pss_aov)
effectsize::eta_squared(rws_aov)
```

```{r}
# Table the results
swls_results <- nice(swls_aov)
pss_results  <- nice(pss_aov)
rws_results  <- nice(rws_aov)

swls_results
pss_results
rws_results
```

# Post Hoc Comparisons
```{r}
# SWLS Post Hoc Comparisons
# Estimated marginal means
swls_emm <- emmeans(swls_aov, ~ group | time)

# Compare groups at each time point
pairs(swls_emm, adjust = "bonferroni")

# Compare time points within each group
swls_time <- emmeans(swls_aov, ~ time | group)
pairs(swls_time, adjust = "bonferroni")
```

```{r}
# PSS Post Hoc Comparisons
# Estimated marginal means
pss_emm <- emmeans(pss_aov, ~ group | time)

# Compare groups at each time point
pairs(pss_emm, adjust = "bonferroni")

# Compare time points within each group
pss_time <- emmeans(pss_aov, ~ time | group)
pairs(pss_time, adjust = "bonferroni")
```

```{r}
# RWS Post Hoc Comparisons
# Estimated marginal means
rws_emm <- emmeans(rws_aov, ~ group | time)

# Compare groups at each time point
pairs(rws_emm, adjust = "bonferroni")

# Compare time points within each group
rws_time <- emmeans(rws_aov, ~ time | group)
pairs(rws_time, adjust = "bonferroni")
```

```{r}
# Summary Table
swls_group_results <- summary(pairs(swls_emm))
swls_time_results <- summary(pairs(swls_time))

pss_group_results <- summary(pairs(pss_emm))
pss_time_results <- summary(pairs(pss_time))

rws_group_results <- summary(pairs(rws_emm))
rws_time_results <- summary(pairs(rws_time))
```

```{r}
summary(swls_emm)
summary(pss_emm)
summary(rws_emm)
```

# Visualizations
```{r}
# Summary Statistics for the Graphs
swls_summary <- data_long %>%
  group_by(group, time) %>%
  summarise(
    Mean = mean(swls, na.rm = TRUE),
    SD = sd(swls, na.rm = TRUE),
    N = sum(!is.na(swls)),
    SE = SD / sqrt(N),
    .groups = "drop")

swls_summary

pss_summary <- pss_long %>%
  group_by(group, time) %>%
  summarise(
    Mean = mean(pss, na.rm = TRUE),
    SD = sd(pss, na.rm = TRUE),
    N = sum(!is.na(pss)),
    SE = SD / sqrt(N),
    .groups = "drop")

pss_summary

rws_summary <- rws_long %>%
  group_by(group, time) %>%
  summarise(
    Mean = mean(rws, na.rm = TRUE),
    SD = sd(rws, na.rm = TRUE),
    N = sum(!is.na(rws)),
    SE = SD / sqrt(N),
    .groups = "drop")

rws_summary
```

```{r}
# SWLS Figure
ggplot(swls_summary, aes(x = time, y = Mean, group = group, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(
    ymin = Mean - SE,
    ymax = Mean + SE),
    width = .15) +
  labs(
    title = "Satisfaction with Life Across Time",
    x = "Assessment Time",
    y = "Mean SWLS Score",
    color = "Group") +
  theme_minimal(base_size = 14)
ggsave("SWLS_Figure.png", width = 8, height = 6, dpi = 300)
```

```{r}
# PSS Figure
ggplot(pss_summary, aes(x = time, y = Mean, group = group, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(
    ymin = Mean - SE,
    ymax = Mean + SE),
    width = .15) +
  labs(
    title = "Perceived Stress Across Time",
    x = "Assessment Time",
    y = "Mean PSS Score",
    color = "Group") +
  theme_minimal(base_size = 14)
ggsave("PSS_Figure.png", width = 8, height = 6, dpi = 300)
```

```{r}
# RWS Figure
ggplot(rws_summary, aes(x = time, y = Mean, group = group, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(
    ymin = Mean - SE,
    ymax = Mean + SE),
    width = .15) +
  labs(title = "Resilience Across Time",
    x = "Assessment Time",
    y = "Mean RWS Score",
    color = "Group") +
  theme_minimal(base_size = 14)
ggsave("RWS_Figure.png", width = 8, height = 6, dpi = 300)
```

```{r}
# Save the Figures
ggsave("SWLS_Figure.png", width = 8, height = 6, dpi = 300)
ggsave("PSS_Figure.png", width = 8, height = 6, dpi = 300)
ggsave("RWS_Figure.png", width = 8, height = 6, dpi = 300)
```

# Exporting Tables and Figures
```{r}
# Table 1: Participant Characteristics
table1 <- data %>%
  group_by(group) %>%
  summarise(
    Participants = n(),
    Age_Mean = mean(age, na.rm = TRUE),
    Age_SD = sd(age, na.rm = TRUE),
    SWLS_Baseline = mean(swls_b, na.rm = TRUE),
    PSS_Baseline = mean(pss_b, na.rm = TRUE),
    RWS_Baseline = mean(rws_b, na.rm = TRUE))

table1 %>%
  gt() %>%
  tab_header(title = md("**Table 1**"), subtitle = "Participant Characteristics by Group") %>%
  fmt_number(columns = where(is.numeric), decimals = 2)
```

```{r}
# Helper Function
mean_sd <- function(x){
  sprintf("%.2f ± %.2f", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))}
```

```{r}
# Table 2
table2 <- data %>%
  group_by(group) %>%
  summarise(
    SWLS_Baseline = mean_sd(swls_b),
    SWLS_Day10 = mean_sd(swls_d10),
    SWLS_Day30 = mean_sd(swls_d30),
    PSS_Baseline = mean_sd(pss_b),
    PSS_Day10 = mean_sd(pss_d10),
    PSS_Day30 = mean_sd(pss_d30),
    RWS_Baseline = mean_sd(rws_b),
    RWS_Day10 = mean_sd(rws_d10),
    RWS_Day30 = mean_sd(rws_d30))

table2 %>%
  gt() %>%
  tab_header(title = md("**Table 2**"), subtitle = "Means (± SD) Across Assessment Time Points")
```

```{r}
# Table 3
swls_results$Outcome <- "SWLS"
pss_results$Outcome <- "PSS"
rws_results$Outcome <- "RWS"

anova_table <- bind_rows(swls_results, pss_results, rws_results)

anova_table %>%
  select(Outcome, Effect, df, F, p.value, ges) %>%
  gt() %>%
  tab_header(title = md("**Table 3**"), subtitle = "Mixed Repeated-Measures ANOVA Results") %>%
  fmt_number(columns = c(F, ges), decimals = 2)
```

```{r}
library(rempsyc)
library(flextable)

# 1. Convert your raw data into nice_table objects
table1_formatted <- nice_table(table1)
table2_formatted <- nice_table(table2)

# 2. Save them as images
save_as_image(table1_formatted, path = "Table1.png")
save_as_image(table2_formatted, path = "Table2.png")

library(gt)

# Convert using gt() instead of nice_table()
anova_formatted <- gt(anova_table)

# Save it normally
gtsave(anova_formatted, "Table3.png")
```

```{r}
# Descriptive Table
library(gtsummary)
library(gt)
library(dplyr)

# 1. Generate the demographic table
demo_table <- data %>%
  # Select only the variables you want to display and the splitting condition
  select(age, gender, group) %>%
  
  # Build the summary table split by your randomized condition
  tbl_summary(
    by = group, # Splits columns by your randomized groups
    missing = "ifany", # Automatically adds a row for missing data if any exist
    missing_text = "Missing", # Label for your missing data rows
    
    # Specify custom formatting for age and gender
    statistic = list(
      age ~ "{mean} ({sd}) [range: {min}, {max}]", # Mean, SD, Min, Max
      gender ~ "{n} ({p}%)"                        # Count and Percentage
    ),
    
    # Provide clean labels for the table rows
    label = list(
      age ~ "Age (Years)",
      gender ~ "Gender"
    )
  ) %>%
  
  # Add an "Overall" column to show the entire sample combined
  add_overall(last = FALSE) %>% 
  
  # Bold the variable names for a clean, professional look
  bold_labels()

# 2. View your table in RStudio
demo_table

```

```{r}
# Convert the summary to a gt object and save as PNG
demo_table %>%
  as_gt() %>%
  gtsave("Demographics_Table.png")
```

