---
title: "HUDM - 5123 Final Project"
author: "Jon Zhang"
date: "2026-06-18"
output: html_document
---
###############################################################
# HUDM 5123 Final Project
#
# Replication of:
# Champion, Economides, & Chandler (2018)
#
# The efficacy of a brief app-based mindfulness intervention
# on psychosocial outcomes in healthy adults
#
# Author: Jonathan Zhang & Luyi Xu
# Date: 07/01/2026
###############################################################
# Section 1: Load Packages
```{r}
# Packages
library(httr)
library(readxl)
library(tidyverse)
library(janitor)
library(psych)
library(afex)
library(emmeans)
library(effectsize)
library(car)
library(rstatix)
```


```{r}
# 2. Define the RAW GitHub URL (Must end in .xlsx or .xls)
github_raw_url <- "https://github.com/ZeddyCraft/HUDM-5123-Final-Project/raw/refs/heads/main/Raw%20Data.xlsx"

# 3. Create a temporary local file path
temp_file <- tempfile(fileext = ".xlsx")

# 4. Download the file securely
GET(github_raw_url, write_disk(temp_file, overwrite = TRUE))

# 5. Read the Excel file into R
data <- read_excel(temp_file)

# View the first few rows
head(data)
```

# Section 2: Import and Clean the Dataset
```{r}
# Clean variable names
data <- clean_names(data)

# Remove completely empty rows
data <- data %>%
  remove_empty(which = "rows")

# Create Participant ID
data <- data %>%
  mutate(id = 1:n())

# Move ID to the first column
data <- data %>%
  relocate(id)

glimpse(data)
```
#@ 2.5 Rename variables
```{r}
data <- data %>%
  rename(
    group = group_a_mindfulness_b_wait_list,
    gender = gender_1_male_2_female,
    
    swls_b = swls_baseline,
    swls_d10 = swls_day_10,
    swls_d30 = swls_day_30,
    
    pss_b = pss_baseline,
    pss_d10 = pss_day_10,
    pss_d30 = pss_day_30,
    
    rws_b = rws_baseline,
    rws_d10 = rws_day_10,
    rws_d30 = rws_day_30,

    sessions_10 = number_of_days_headspace_used_between_baseline_to_day_10,
    sessions_30 = number_of_days_headspace_used_between_day_10_and_day_30,

    enjoy_10 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_10_1_not_at_all_to_7_extremely,
    enjoy_30 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_30_1_not_at_all_to_7_extremely,

    difficulty_10 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_10_1_very_difficult_to_7_very_easy,
    difficulty_30 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_30_1_very_difficult_to_7_very_easy
  )

data <- data %>%
  mutate(
    group = factor(group,
                   levels = c("A", "B"),
                   labels = c("Mindfulness", "Wait-list")),

    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )
```

# Section 3: Descriptive Statistics
```{r}
# Number of participants in each group
table(data$group)

# Percentage in each group
prop.table(table(data$group)) * 100

# Summary of age
describe(data$age)

# Summary by group
data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  )

# Gender distribution
table(data$group, data$gender)

# Gender percentages within each group
prop.table(table(data$group,
                 data$gender), margin = 1) * 100
```

## Section 3.1: Means and Standard Deviations
```{r}
data %>%
  group_by(group) %>%
  summarise(
    # Satisfaction with Life
    SWLS_B_Mean = mean(swls_b, na.rm = TRUE),
    SWLS_B_SD   = sd(swls_b, na.rm = TRUE),

    SWLS_D10_Mean = mean(swls_d10, na.rm = TRUE),
    SWLS_D10_SD   = sd(swls_d10, na.rm = TRUE),

    SWLS_D30_Mean = mean(swls_d30, na.rm = TRUE),
    SWLS_D30_SD   = sd(swls_d30, na.rm = TRUE),

    # Perceived Stress
    PSS_B_Mean = mean(pss_b, na.rm = TRUE),
    PSS_B_SD   = sd(pss_b, na.rm = TRUE),

    PSS_D10_Mean = mean(pss_d10, na.rm = TRUE),
    PSS_D10_SD   = sd(pss_d10, na.rm = TRUE),

    PSS_D30_Mean = mean(pss_d30, na.rm = TRUE),
    PSS_D30_SD   = sd(pss_d30, na.rm = TRUE),

    # Resilience
    RWS_B_Mean = mean(rws_b, na.rm = TRUE),
    RWS_B_SD   = sd(rws_b, na.rm = TRUE),

    RWS_D10_Mean = mean(rws_d10, na.rm = TRUE),
    RWS_D10_SD   = sd(rws_d10, na.rm = TRUE),

    RWS_D30_Mean = mean(rws_d30, na.rm = TRUE),
    RWS_D30_SD   = sd(rws_d30, na.rm = TRUE)
  )
```

# Section 4: Reshape the Data (wide to Long)
```{r}
# Satisfaction With Life Scale
data_long <- data %>%
  select(id, group, swls_b, swls_d10, swls_d30) %>%
  pivot_longer(cols = starts_with("swls"), names_to = "time", values_to = "swls") %>%
  mutate(time = case_when(
         time == "swls_b"   ~ "Baseline",
         time == "swls_d10" ~ "Day 10",
         time == "swls_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Perceived Stress Scale (PSS)
pss_long <- data %>%
  select(id, group, pss_b, pss_d10, pss_d30) %>%
  pivot_longer(cols = starts_with("pss"), names_to = "time", values_to = "pss") %>%
  mutate(time = case_when(
         time == "pss_b" ~ "Baseline",
         time == "pss_d10" ~ "Day 10",
         time == "pss_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Resilience Scale (RWS)
rws_long <- data %>%
  select(id, group, rws_b, rws_d10, rws_d30) %>%
  pivot_longer(cols = starts_with("rws"), names_to = "time", values_to = "rws") %>%
  mutate(time = case_when(
         time == "rws_b" ~ "Baseline",
         time == "rws_d10" ~ "Day 10",
         time == "rws_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")))

# Check the dataset
head(rws_long)
glimpse(rws_long)
```

# Section 5 Assumption Checks
## Section 5.1: Missing Data
```{r}
# Missing Data
## Missing values in each long dataset
colSums(is.na(data_long))
colSums(is.na(pss_long))
colSums(is.na(rws_long))

## Total missing observations
sum(is.na(data_long$swls))
sum(is.na(pss_long$pss))
sum(is.na(rws_long$rws))
```
```{r}
# Outliers (by Boxplots)
## SWLS
ggplot(data_long, aes(x = time, y = swls,fill = group)) +
  geom_boxplot() +
  labs(title = "SWLS by Time and Group", x = "Time", y = "SWLS Score") +
  theme_minimal()

## PSS
ggplot(pss_long, aes(x = time, y = pss, fill = group)) +
  geom_boxplot() +
  labs(title = "PSS by Time and Group", x = "Time", y = "PSS Score") +
  theme_minimal()

## RWS
ggplot(rws_long, aes(x = time, y = rws, fill = group)) +
  geom_boxplot() +
  labs(title = "RWS by Time and Group", x = "Time", y = "RWS Score") +
  theme_minimal()
```

```{r}
# Outliers (by Outlier Detection)
data_long %>%
  group_by(group, time) %>%
  identify_outliers(swls)

pss_long %>%
  group_by(group, time) %>%
  identify_outliers(pss)

rws_long %>%
  group_by(group, time) %>%
  identify_outliers(rws)
```

```{r}
# Normality
## SWLS
data_long %>%
  group_by(group, time) %>%
  shapiro_test(swls)

## PSS
pss_long %>%
  group_by(group, time) %>%
  shapiro_test(pss)

## RWS
rws_long %>%
  group_by(group, time) %>%
  shapiro_test(rws)
```

```{r}
## QQ Plots
## SWLS
ggplot(data_long, aes(sample = swls)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()

## PSS
ggplot(pss_long, aes(sample = pss)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()

## RWS
ggplot(rws_long, aes(sample = rws)) +
  stat_qq() +
  stat_qq_line() +
  facet_grid(group ~ time) +
  theme_minimal()
```

```{r}
# Homogeneity of Variance (Levene's Test) 
## SWLS
leveneTest(swls ~ group * time, data = data_long)

## PSS
leveneTest(pss ~ group * time,data = pss_long)

## RWS
leveneTest(rws ~ group * time, data = rws_long)
```
# Mixed Repeated-Measures ANOVA
```{r}
###############################################################
# Section 6.1: Mixed ANOVA - Satisfaction With Life (SWLS)
###############################################################

swls_aov <- aov_ez(
  id = "id",
  dv = "swls",
  between = "group",
  within = "time",
  data = data_long,
  type = 3
)

# Display ANOVA table
summary(swls_aov)

# Detailed ANOVA table
nice(swls_aov)
```

