---
title: "HUDM 5123 Final Project — Revised Analysis"
author: "Luyi Xu & Jonathan Zhang"
date: "2026-06-28"
output: html_document
---

<!--
########################################################################
# luyi note (READ ME FIRST — what I changed and why)
#
# Hi Jonathan — thank you for building this out, your pipeline is doing
# almost all the work and I kept nearly all of it. The one thing I
# changed is the MODEL ENGINE, because our Methods section promises a
# likelihood-based linear mixed-effects model (LMM) and the original
# paper's PRIMARY analysis was an LMM too — not the repeated-measures
# ANOVA. Here is the short version of the edits:
#
# 1. aov_ez() (afex)  ->  lmer() (lmerTest).
#    Reason: aov_ez does LISTWISE DELETION. It drops the 12 participants
#    who are missing any one occasion (9 of them from the MINDFULNESS arm),
#    so it quietly throws away 12 observed PSS scores AND replicates only
#    the paper's *secondary* complete-case ANOVA. The LMM keeps everyone
#    with >= 1 observed score (available-case under MAR) and replicates the
#    paper's *primary* (ITT) analysis. This also makes Results match the
#    Methods we already wrote.
#
# 2. PSS is now the CONFIRMATORY outcome with two models:
#       (a) replication LMM, numeric time 0/10/30  (matches Champion et al.)
#       (b) extension LMM, categorical time         (our early-vs-late extension)
#    SWLS and RWS are still here but clearly LABELED EXPLORATORY.
#
# 3. lmerTest (not base lme4) is loaded so we get the Satterthwaite
#    degrees of freedom our Methods section promises.
#
# 4. Planned contrasts now come from emmeans on the categorical LMM, with
#    Holm adjustment. The Baseline->Day 30 difference-in-change is PRIMARY.
#
# 5. The PSS figure now plots MODEL-estimated means with 95% CIs (our plan
#    says model-estimated means, not raw cell means).
#
# 6. Your aov_ez ANOVA is NOT deleted — I moved it to the bottom as a
#    complete-case SENSITIVITY check, which is exactly where it belongs.
#
# Search this file for "luyi note:" to see each change in context.
########################################################################
-->

# Load the Packages
```{r setup, message=FALSE, warning=FALSE}
# Packages
library(httr)
library(readxl)     # for reading .xlsx files
library(tidyverse)  # dplyr, tidyr, ggplot2, etc.
library(janitor)
library(psych)
library(emmeans)    # estimated marginal means + planned contrasts
library(effectsize)
library(car)
library(rstatix)

# luyi note: NEW — lmerTest::lmer() is our primary model engine now. It is the
# same as lme4::lmer() but adds Satterthwaite degrees of freedom + p-values,
# which is what our Methods section commits us to. Loading lmerTest also makes
# emmeans use lmer-aware dfs automatically.
library(lmerTest)

# luyi note: afex is KEPT only for the optional complete-case RM-ANOVA
# sensitivity check at the very bottom. It is no longer our main model.
library(afex)

# For tables
library(gt)
library(gtsummary)

# luyi note: force Satterthwaite dfs in emmeans so the contrast dfs match the
# Methods text (otherwise emmeans may default to Kenward-Roger if pbkrtest is
# installed). Either is defensible; we just picked one and documented it.
emmeans::emm_options(lmer.df = "satterthwaite")
```

# Uploading the Dataset
```{r}
# luyi note: the GitHub download is fine for us, but for the GRADED submission
# the script must also run from the attached local file (the rubric wants
# replicable results from what we hand in). I left your download in place and
# added a commented local-read fallback right below it. Use whichever path the
# grader has, and make sure the submitted .Rmd sits next to "Raw_Data.xlsx".

github_raw_url <- "https://github.com/ZeddyCraft/HUDM-5123-Final-Project/raw/refs/heads/main/Raw%20Data.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
GET(github_raw_url, write_disk(temp_file, overwrite = TRUE))
data <- read_excel(temp_file)

# luyi note: local fallback for the submission (uncomment this and comment the
# three lines above if reading the attached file directly):
# data <- read_excel("Raw_Data.xlsx")

head(data)
```

# Importing and Cleaning the Dataset
```{r}
data <- clean_names(data)
data <- data %>% remove_empty(which = "rows")
data <- data %>% mutate(id = 1:n())
data <- data %>% relocate(id)
glimpse(data)
```

# Rename variables
```{r}
data <- data %>%
  rename(
    group = group_a_mindfulness_b_wait_list,
    gender = gender_1_male_2_female,

    swls_b = swls_baseline,
    swls_d10 = swls_day_10,
    swls_d30 = swls_day_30,

    pss_b = pss_baseline,
    pss_d10 = pss_day_10,
    pss_d30 = pss_day_30,

    rws_b = rws_baseline,
    rws_d10 = rws_day_10,
    rws_d30 = rws_day_30,

    sessions_10 = number_of_days_headspace_used_between_baseline_to_day_10,
    sessions_30 = number_of_days_headspace_used_between_day_10_and_day_30,

    enjoy_10 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_10_1_not_at_all_to_7_extremely,
    enjoy_30 = to_what_to_extent_are_you_finding_completing_the_activities_enjoyable_day_30_1_not_at_all_to_7_extremely,

    difficulty_10 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_10_1_very_difficult_to_7_very_easy,
    difficulty_30 = to_what_extent_are_you_finding_the_activity_difficult_to_complete_day_30_1_very_difficult_to_7_very_easy
  )

data <- data %>%
  mutate(
    group = factor(group, levels = c("A", "B"),
                   labels = c("Mindfulness", "Wait-list")),
    gender = factor(gender, levels = c(1, 2),
                    labels = c("Male", "Female"))
  )
```

# Descriptive Statistics
```{r}
table(data$group)
prop.table(table(data$group)) * 100
describe(data$age)

data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  )

table(data$group, data$gender)
prop.table(table(data$group, data$gender), margin = 1) * 100
```

```{r}
# Means and Standard Deviations (kept exactly as you had it — good descriptive table)
data %>%
  group_by(group) %>%
  summarise(
    SWLS_B_Mean = mean(swls_b, na.rm = TRUE), SWLS_B_SD = sd(swls_b, na.rm = TRUE),
    SWLS_D10_Mean = mean(swls_d10, na.rm = TRUE), SWLS_D10_SD = sd(swls_d10, na.rm = TRUE),
    SWLS_D30_Mean = mean(swls_d30, na.rm = TRUE), SWLS_D30_SD = sd(swls_d30, na.rm = TRUE),
    PSS_B_Mean = mean(pss_b, na.rm = TRUE), PSS_B_SD = sd(pss_b, na.rm = TRUE),
    PSS_D10_Mean = mean(pss_d10, na.rm = TRUE), PSS_D10_SD = sd(pss_d10, na.rm = TRUE),
    PSS_D30_Mean = mean(pss_d30, na.rm = TRUE), PSS_D30_SD = sd(pss_d30, na.rm = TRUE),
    RWS_B_Mean = mean(rws_b, na.rm = TRUE), RWS_B_SD = sd(rws_b, na.rm = TRUE),
    RWS_D10_Mean = mean(rws_d10, na.rm = TRUE), RWS_D10_SD = sd(rws_d10, na.rm = TRUE),
    RWS_D30_Mean = mean(rws_d30, na.rm = TRUE), RWS_D30_SD = sd(rws_d30, na.rm = TRUE)
  )
```

# Reshaping the Data (wide to long)
```{r}
# luyi note: I kept your three long datasets, but added a NUMERIC time column
# (time_num = 0/10/30) to each. The replication LMM needs numeric time to match
# the paper; the extension LMM uses the factor "time". Both live in the same
# long frame now. (I left the SWLS frame named data_long so your downstream
# SWLS code still runs — though a clearer name would be swls_long.)

# Satisfaction With Life Scale (SWLS)
data_long <- data %>%
  select(id, group, swls_b, swls_d10, swls_d30) %>%
  pivot_longer(cols = starts_with("swls"), names_to = "time", values_to = "swls") %>%
  mutate(
    time = case_when(time == "swls_b" ~ "Baseline",
                     time == "swls_d10" ~ "Day 10",
                     time == "swls_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")),
    time_num = recode(as.character(time), "Baseline" = 0, "Day 10" = 10, "Day 30" = 30)  # luyi note: numeric time
  )

# Perceived Stress Scale (PSS) — CONFIRMATORY outcome
pss_long <- data %>%
  select(id, group, pss_b, pss_d10, pss_d30) %>%
  pivot_longer(cols = starts_with("pss"), names_to = "time", values_to = "pss") %>%
  mutate(
    time = case_when(time == "pss_b" ~ "Baseline",
                     time == "pss_d10" ~ "Day 10",
                     time == "pss_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")),
    time_num = recode(as.character(time), "Baseline" = 0, "Day 10" = 10, "Day 30" = 30)
  )

# Resilience Scale (RWS)
rws_long <- data %>%
  select(id, group, rws_b, rws_d10, rws_d30) %>%
  pivot_longer(cols = starts_with("rws"), names_to = "time", values_to = "rws") %>%
  mutate(
    time = case_when(time == "rws_b" ~ "Baseline",
                     time == "rws_d10" ~ "Day 10",
                     time == "rws_d30" ~ "Day 30"),
    time = factor(time, levels = c("Baseline", "Day 10", "Day 30")),
    time_num = recode(as.character(time), "Baseline" = 0, "Day 10" = 10, "Day 30" = 30)
  )

glimpse(pss_long)
```

# Assumption / Data Screening (descriptive)
```{r}
# luyi note: I kept all of your screening (missingness, boxplots, outliers,
# Shapiro, Q-Q, Levene). For the LMM these are DESCRIPTIVE screening rather
# than formal model assumptions — the formal LMM diagnostics are residual-based
# and I added them right after the PSS models. Sphericity is NOT required for
# the LMM (only for RM-ANOVA), so we no longer lean on it for the main analysis.

# Missing values
colSums(is.na(data_long))
colSums(is.na(pss_long))
colSums(is.na(rws_long))

# luyi note: this confirms the key fact for our missing-data paragraph —
# PSS has 24 missing CELLS (2 baseline / 10 day-10 / 12 day-30) and 62 of 74
# participants are complete on all three. Listwise deletion would drop the
# other 12 (9 mindfulness, 3 wait-list); our LMM keeps their observed scores.
n_complete_pss <- pss_long %>% group_by(id) %>%
  summarise(n_obs = sum(!is.na(pss)), .groups = "drop") %>%
  count(complete = n_obs == 3)
n_complete_pss
```

```{r}
# Boxplots
ggplot(data_long, aes(x = time, y = swls, fill = group)) + geom_boxplot() +
  labs(title = "SWLS by Time and Group", x = "Time", y = "SWLS Score") + theme_minimal()
ggplot(pss_long, aes(x = time, y = pss, fill = group)) + geom_boxplot() +
  labs(title = "PSS by Time and Group", x = "Time", y = "PSS Score") + theme_minimal()
ggplot(rws_long, aes(x = time, y = rws, fill = group)) + geom_boxplot() +
  labs(title = "RWS by Time and Group", x = "Time", y = "RWS Score") + theme_minimal()
```

```{r}
# Outlier detection (descriptive)
data_long %>% group_by(group, time) %>% identify_outliers(swls)
pss_long  %>% group_by(group, time) %>% identify_outliers(pss)
rws_long  %>% group_by(group, time) %>% identify_outliers(rws)
```

```{r}
# Normality per cell (descriptive)
data_long %>% group_by(group, time) %>% shapiro_test(swls)
pss_long  %>% group_by(group, time) %>% shapiro_test(pss)
rws_long  %>% group_by(group, time) %>% shapiro_test(rws)
```

```{r}
# Levene (descriptive). luyi note: drop NA rows so Levene runs cleanly.
leveneTest(swls ~ group * time, data = data_long %>% drop_na(swls))
leveneTest(pss  ~ group * time, data = pss_long  %>% drop_na(pss))
leveneTest(rws  ~ group * time, data = rws_long  %>% drop_na(rws))
```

# =====================================================================
# CONFIRMATORY ANALYSIS — PERCEIVED STRESS (PSS)
# =====================================================================

## PSS Model 1 — Replication LMM (numeric time, matches Champion et al.)
```{r}
# luyi note: this is the REPLICATION model. Champion et al. coded time as
# 0/10/30, group as 1/0, with random intercepts AND slopes. To make our
# interaction coefficient carry the SAME SIGN as their reported beta = -0.28,
# I relevel so Wait-list is the reference group (so "Mindfulness" is the 1).
pss_long <- pss_long %>% mutate(group_rep = relevel(group, ref = "Wait-list"))

pss_rep <- lmer(pss ~ group_rep * time_num + (1 + time_num | id),
                data = pss_long, REML = TRUE)

# luyi note: with only 3 timepoints a random slope can give a singular fit.
# Check isSingular(); if TRUE, fall back in this order (uncomment as needed):
# pss_rep <- lmer(pss ~ group_rep * time_num + (1 + time_num || id), data = pss_long, REML = TRUE)
# pss_rep <- lmer(pss ~ group_rep * time_num + (1 | id),             data = pss_long, REML = TRUE)
isSingular(pss_rep)

summary(pss_rep)                      # group_rep:time_num row ~ the paper's interaction beta
confint(pss_rep, method = "Wald")     # 95% CI for the interaction
```

## PSS Model 2 — Extension LMM (categorical time; PRIMARY for our contrasts)
```{r}
# luyi note: this is our EXTENSION + the model we use for all planned contrasts.
# Categorical time lets baseline->Day 10 (early) change be separated from
# Day 10->Day 30 (late) change instead of forcing one linear slope.
pss_ext <- lmer(pss ~ group * time + (1 | id), data = pss_long, REML = TRUE)
isSingular(pss_ext)

# Satterthwaite F-tests; the group:time row is the OMNIBUS interaction test.
anova(pss_ext)
summary(pss_ext)
```

## PSS — does the extension improve on the linear form? (clean nested LRT)
```{r}
# luyi note: linear-in-time is a special case of (nested within) categorical
# time, so this is a clean test of whether the trajectory departs from linear.
# Both refit with ML and SAME random structure so only the fixed time form
# differs (this is the honest comparison; comparing the random-slope model to
# the categorical model would confound fixed + random differences).
pss_lin_ri <- lmer(pss ~ group * time_num + (1 | id), data = pss_long, REML = FALSE)
pss_cat_ri <- lmer(pss ~ group * time     + (1 | id), data = pss_long, REML = FALSE)
anova(pss_lin_ri, pss_cat_ri)   # AIC/BIC + LRT for departure from linearity
```

## PSS — Planned contrasts (Holm-adjusted; Baseline->Day 30 is PRIMARY)
```{r}
# luyi note: this replaces your pairs(..., adjust="bonferroni"). It computes the
# BETWEEN-CONDITION DIFFERENCE-IN-CHANGE (Mindfulness minus Wait-list) for
# baseline->Day 10 and baseline->Day 30. The Day 30 row is our PRIMARY estimand.
emm_pss <- emmeans(pss_ext, ~ time * group)

did_pss <- contrast(
  emm_pss,
  interaction = c("trt.vs.ctrl1", "pairwise"),  # time vs Baseline; group MM vs WL
  adjust = "holm"
)
did_pss
confint(did_pss)
```

```{r}
# Extension narrative: WITHIN-group early vs late change (consecutive contrasts)
# luyi note: this is what makes the categorical model worth doing — it shows
# whether each group changed early (Day10-Baseline) vs late (Day30-Day10).
emm_pss_tg <- emmeans(pss_ext, ~ time | group)
within_change_pss <- contrast(emm_pss_tg, "consec", adjust = "holm")
within_change_pss
```

```{r}
# Supplementary (descriptive): group difference at each occasion.
# luyi note: keep this as SUPPLEMENTARY only — our Methods says the headline
# table is the difference-in-change contrasts, not every pairwise cell.
emm_pss_gt <- emmeans(pss_ext, ~ group | time)
pairs(emm_pss_gt, adjust = "holm")
```

## PSS — Standardized effect size for the primary (Day 30) contrast
```{r}
# luyi note: matches the paper's between-group d = (difference in change) /
# (pooled baseline SD). Champion et al. reported d ~ 1.42-1.53 for PSS; ours
# will be close (sign is negative here = stress reduced in mindfulness vs WL).
base_sd_pss <- pss_long %>% filter(time == "Baseline") %>%
  summarise(sd = sd(pss, na.rm = TRUE)) %>% pull(sd)

did_df <- as.data.frame(did_pss)
# the Day 30 - Baseline row is the second row of did_pss:
d30_estimate <- did_df$estimate[grepl("Day 30", did_df[[1]])]
d_primary_pss <- d30_estimate / base_sd_pss
cat("Pooled baseline SD (PSS):", round(base_sd_pss, 2),
    "| Primary Day-30 standardized effect:", round(d_primary_pss, 2), "\n")
```

## PSS — LMM diagnostics (these are the REAL assumption checks for our model)
```{r}
# luyi note: NEW. For an LMM the assumptions are about residuals and random
# effects, not sphericity. Report these for the final PSS extension model.
plot(pss_ext)                                   # residuals vs fitted (constant variance)
qqnorm(resid(pss_ext)); qqline(resid(pss_ext))  # residual normality
re_pss <- ranef(pss_ext)$id[, 1]
qqnorm(re_pss); qqline(re_pss)                   # random-intercept normality
```

## PSS — Model-estimated profile plot (95% CI)
```{r}
# luyi note: our plan calls for MODEL-estimated means, so this plots emmeans
# from the LMM with 95% CIs (replacing the raw-mean + SE version).
pss_emm_plot <- as.data.frame(emmeans(pss_ext, ~ group * time))

ggplot(pss_emm_plot, aes(x = time, y = emmean, group = group, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .15) +
  labs(title = "Model-Estimated Perceived Stress Across Time (LMM)",
       x = "Assessment Time", y = "Estimated Mean PSS (95% CI)", color = "Group") +
  theme_minimal(base_size = 14)
ggsave("PSS_Figure.png", width = 8, height = 6, dpi = 300)
```

# =====================================================================
# EXPLORATORY OUTCOMES — SWLS and RWS
# luyi note: SAME two LMMs as PSS, but these are EXPLORATORY (not confirmatory).
# Per our Methods, SWLS/RWS are reported as exploratory so we don't inflate the
# number of confirmatory tests. I used a small helper to avoid duplicating the
# PSS code; the PSS block above is the canonical, fully-commented template.
# =====================================================================
```{r}
fit_lmm_pair <- function(df, dv){
  f_lin <- as.formula(paste0(dv, " ~ group * time_num + (1 | id)"))
  f_cat <- as.formula(paste0(dv, " ~ group * time     + (1 | id)"))
  list(
    rep = lmer(f_lin, data = df, REML = TRUE),   # numeric-time replication form
    ext = lmer(f_cat, data = df, REML = TRUE)    # categorical-time extension
  )
}

explore_outcome <- function(df, dv){
  fits <- fit_lmm_pair(df, dv)
  cat("\n================ EXPLORATORY:", toupper(dv), "================\n")
  print(anova(fits$ext))                                  # omnibus interaction
  emm <- emmeans(fits$ext, ~ time * group)
  did <- contrast(emm, interaction = c("trt.vs.ctrl1", "pairwise"), adjust = "holm")
  print(did)                                              # difference-in-change
  print(confint(did))
  plot_df <- as.data.frame(emmeans(fits$ext, ~ group * time))
  p <- ggplot(plot_df, aes(time, emmean, group = group, color = group)) +
    geom_line(linewidth = 1.2) + geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = .15) +
    labs(title = paste0("Model-Estimated ", toupper(dv),
                        " Across Time (LMM, exploratory)"),
         x = "Assessment Time", y = paste0("Estimated Mean ", toupper(dv), " (95% CI)"),
         color = "Group") +
    theme_minimal(base_size = 14)
  print(p)
  invisible(fits)
}

swls_fits <- explore_outcome(data_long, "swls")  # data_long holds SWLS
rws_fits  <- explore_outcome(rws_long,  "rws")
```

# =====================================================================
# SENSITIVITY ANALYSIS — complete-case
# =====================================================================
```{r}
# luyi note: your aov_ez ANOVA is NOT thrown away — it becomes a complete-case
# sensitivity check (PSS only). We also refit the SAME categorical LMM on the
# 62 completers and confirm the interaction conclusion is unchanged.

# (a) complete-case LMM
pss_complete <- pss_long %>% group_by(id) %>%
  filter(sum(!is.na(pss)) == 3) %>% ungroup()
pss_ext_cc <- lmer(pss ~ group * time + (1 | id), data = pss_complete, REML = TRUE)
anova(pss_ext_cc)

# (b) complete-case RM-ANOVA (your original engine, repurposed as sensitivity)
# luyi note: afex needs complete cases; this is roughly the paper's *secondary*
# analysis. Useful as a cross-check, NOT as the primary result.
pss_aov_cc <- aov_ez(id = "id", dv = "pss", between = "group", within = "time",
                     data = pss_complete, type = 3)
nice(pss_aov_cc)
```

# =====================================================================
# RESULTS TABLES (APA-style export)
# =====================================================================
```{r}
# Table 1 — participant characteristics (kept from your script)
table1 <- data %>%
  group_by(group) %>%
  summarise(
    Participants = n(),
    Age_Mean = mean(age, na.rm = TRUE),
    Age_SD   = sd(age, na.rm = TRUE),
    PSS_Baseline = mean(pss_b, na.rm = TRUE),
    SWLS_Baseline = mean(swls_b, na.rm = TRUE),
    RWS_Baseline = mean(rws_b, na.rm = TRUE)
  )
table1 %>%
  gt() %>%
  tab_header(title = md("**Table 1**"), subtitle = "Participant Characteristics by Group") %>%
  fmt_number(columns = where(is.numeric), decimals = 2)
```

```{r}
# Table 2 — PSS confirmatory results: omnibus interaction + planned contrasts
# luyi note: this replaces the ANOVA results table. It reports the LMM omnibus
# interaction and the Holm-adjusted difference-in-change contrasts (the numbers
# our Results section should cite).
pss_omnibus <- as.data.frame(anova(pss_ext)) %>%
  rownames_to_column("Effect") %>%
  filter(Effect == "group:time")

pss_contrasts_tbl <- as.data.frame(did_pss) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

gt(pss_contrasts_tbl) %>%
  tab_header(title = md("**Table 2**"),
             subtitle = "PSS difference-in-change contrasts (Mindfulness − Wait-list), Holm-adjusted")

pss_omnibus  # report F, df, p for the group x time interaction in text/table
```

```{r}
# Demographics table (kept from your script)
demo_table <- data %>%
  select(age, gender, group) %>%
  tbl_summary(
    by = group, missing = "ifany", missing_text = "Missing",
    statistic = list(age ~ "{mean} ({sd}) [range: {min}, {max}]",
                     gender ~ "{n} ({p}%)"),
    label = list(age ~ "Age (Years)", gender ~ "Gender")
  ) %>%
  add_overall(last = FALSE) %>%
  bold_labels()
demo_table
```

<!--
########################################################################
# luyi note (handoff checklist for you, Jonathan):
#  - Run top-to-bottom once and check isSingular() on pss_rep. If TRUE, switch
#    to the "||" or random-intercept fallback I left commented, and tell me
#    which one you kept so I can write one sentence about it in Methods.
#  - The PSS group:time interaction (anova(pss_ext)) is our headline number.
#    The did_pss "Day 30" row is the PRIMARY effect; its CI + d go in the text.
#  - SWLS/RWS results print from explore_outcome() — label them EXPLORATORY in
#    the writeup (no strong claims).
#  - I rewrote nothing in your Methods; this script now MATCHES the Methods we
#    already wrote. The Results + Discussion prose still need updating from
#    "three ANOVAs" to "PSS LMM (confirmatory) + SWLS/RWS (exploratory)" — I'll
#    take that pass next.
########################################################################
-->
